Improving Haskell Types with SMT

### 2015-03-14 docs/paper.tex
Typos and clarifications.

For a few years now, there has been a steady push in the Haskell
community to explore and extend the type system, slowly approximating
functionality available in dependently typed languages
\cite{Eisenberg2012,Lindley2013,Eisenberg2014}.  The additional
expressiveness enables Haskell programmers to express more
invariants at compile time, which makes it easier to develop
reliable software, and also makes Haskell a nice language
for embedding domain specific languages \cite{ivory-experience}.
The Haskell compiler is not just a translator, but it also
serves as a tool that analyzes the program, and helps
finds common mistakes early in the development cycle.

@@

We illustrate the utility of the functionality provided by our
algorithm with a few short examples.  A very common example in
this area is to define a family of singleton types that links
type-level natural numbers to run-time constants that represent
them:
\begin{Verbatim}
data UNat :: Nat -> * where
  Zero :: UNat 0
  Succ :: UNat n -> UNat (n + 1)
\end{Verbatim}
Here we've used a unary representation of the natural numbers,
and each member of the family, \Verb"UNat n", has exactly one
inhabitant, namely the natural number \Verb"n" represented in
unary form.  Because we are using a GADT, we can pattern match
on the constructors of the type and gradually learn additional
information about the value being examined.

@@

The rest of the paper is organized as follows:
we start with a brief overview
of SMT solvers form a user's perspective (Section~\ref{smt}); this
section may be skipped by readers familiar with the tools.  Then,
in Section~\ref{GHC}, we introduce the basic concepts of GHC's constraints
solver, which is necessary for putting the rest of the paper in context.
Section~\ref{algorithm} contains the details of the algorithm for
integrating an SMT solver with GHC, and Section~\ref{other} explains
other theories that could be added to GHC using the same technique.
Finally, Section~\ref{modular-typechecker} discusses the possibility
of using the ideas in this paper, and previous work on the
implementation of SMT solver, to engineer a modular constraint solver.

@@

isZero :: E a -> Bool
isZero (EInt x) = x == 0
\end{Verbatim}
When we check the definition of \Verb"isZero", we end up with an implication
constraint like this:
\begin{Verbatim}
(a ~ Int) => (Num a, Eq a)
\end{Verbatim}
Here, \Verb"a" is the type of the pattern variable \Verb"x", the wanted
constraints arise from the use of \Verb"0" and \Verb"(==)" respectively,
and the given constraint is obtain by pattern matching with \Verb"EInt".

@@

\subsection{Type-Checker Plugins}
In this paper we describing a fairly general extension to the constraint
solver, however one could always imagine others too. Indeed, there is
existing work that extends GHC with support for units of measure
\cite{units-of-measure}.  Instead of having many ad-hoc extensions directly
in GHC's constraint solver, we collaborated to define and implement an
API for extending GHC's functionality via {\em type-checker plug-ins}.
At present, we are aware of two main
users of this API---our work, and the work on units of measure. We
hope, however, that this mechanism would be useful to other researchers too,
as it makes it fairly easy to experiment with various extensions to
GHC's constraint solver.

@@

Instead, derived equalities help with type inference, by guiding the
instantiation of {\em unification variables} with types. In general,
it is always sound to instantiate a unification variable with whatever
type we want: doing so may reject valid programs, but it will not accept
invalid ones, because we still need to solve all necessary constraints.
Of course, we don't want to reject valid programs, and this is where derived
equalities help us: since they are implied by the goals and assumptions
together, we know that we are not loosing any generality when instantiating
as suggested by a derived constraint.  So, for example, if we compute a
derived constraint \Verb"x ~ Int", and we have a wanted constraint
\Verb"Eq x", and \Verb"x" is a unification variable, then we may
rewrite \Verb"Eq x" to \Verb"Eq Int", which we would proceed to solve
as usual.

@@

\paragraph{Input.} Currently, the algorithm is implemented as a type-checker
plug-in, and so the input to the algorithm is a collection of constraints
that GHC has determine to be inert. This means that GHC simplified everything
as much as it could, improved using equalities, and basically cannot see
anything else to do.  The inert constraints are presented to the plug-in
in three collections, based on the constraints' provenance: one group contains
the given constraints (i.e., assumptions), one group contains derived
constraints (see Section~\ref{GHC} for details), and one group contains
the wanted constraints, which are the goals that need solving.

@@

\subsection{The Language of Constraints}
Our first task is to identify constraints that are relevant to our solver.
In the current implementation, we consider equalities and inequalities
between a subset of the types of kind \Verb"Nat".   The kind \Verb"Nat" is
inhabited by an infinite family of type-constants:
\begin{Verbatim}
0, 1, 2, .. :: Nat
\end{Verbatim}
These constants may be combined and compared using the following
type-level functions:
\begin{Verbatim}
type family (+)   :: Nat -> Nat -> Nat
type family (*)   :: Nat -> Nat -> Nat
type family (<=?) :: Nat -> Nat -> Bool
\end{Verbatim}
These functions have no user-specified definitions: instead, we extended
the core GHC simplifier with support for forward evaluation on concrete
values, so it will evaluate concrete expressions, such as \Verb"2 + 3".
This simple forward evaluation cleans up the constraints, leaving the more
complex reasoning---involving variables---to the algorithm being presently
described.

@@

The constraints of interest to the algorithm are summarized by the following
grammar:
\begin{align*}
c & = e \overset {\mathbb{N}} \sim e ~|~  e \overset {\mathbb{B}} \sim e \\
e^\mathbb{N} & = \alpha ~|~ n ~|~ e + e ~|~ n * e \\
e^\mathbb{B} & = \alpha ~|~ \mathrm{False} ~|~ \mathrm{True} ~|~ e \leq^? ~e \\
n          & = 0 ~|~ 1 ~|~ \dots
\end{align*}

@@

\paragraph{Natural Numbers.}
We are interested in working with natural number, however, the SMT
decision procedure works with integers.  To work around this mismatch,
we are careful that whenever we declare an SMT variable corresponding
to a type of kind \Verb"Nat", we also add a formula asserting that
the variable is not negative.

@@

The code samples in the rest of this Section use the following
API for interaction with the solver:
\begin{Verbatim}
data Solver :: *
data Result = Sat | Unsat

@@

To check for consistency, we assert all constraints, and ask the SMT
solver if the result is satisfiable.  If this is not the case, then
we report and error, otherwise we proceed with the algorithm.
\begin{Verbatim}
checkConsistent s cs =
  do mapM_ (solverAssert s) cs
     res <- solverCheck s
     return (res == Sat)
\end{Verbatim}

@@

Unfortunately, not all SMT solvers support this feature and CVC4---%
the SMT solver that we used during the development of this work---%
does not have this ability.  To work around this, we implemented
a custom minimization algorithm which work as follows:
\begin{Verbatim}
minimize s yes [] = return yes
minimize s yes mb =
  do solverPush s
     search s yes [] mb

search s yes mb (c : cs) =
  do solverAssert s c
     res <- solverCheck s
     case res of
       Unsat ->
         do solverPop s
            solverAssert s c
            minimize s (c : yes) mb

@@

\paragraph{Improve to Constant.} As we know that the constraints are
consistent, we can ask the SMT solver for a satisfying assignment.
This assignment contains one possible solution to constraints, but
we should not emit an improving equality unless we are sure that
this is the {\em the only} possible solution.  So, if $x = v$ is
in the satisfying assumption, then we try to prove that the currently
asserted constraints imply this fact.  We do this by temporarily asserting
$x \neq v$, and checking for satisfiability.  If the resulting is unsatisfiable,
then we know that $v$ is the only possible value for $x$,
and we can emit the corresponding improving equation.
\begin{Verbatim}
solverProve s p =
  do solverPush s
     solverAssert s (SMT.not p)
     res <- solverCheck s
     solverPop s
     return (res == Unsat)

mustBeK s x v =
  solverPorve s (eq (SMT.const x) (value v))
\end{Verbatim}
This process takes $O(n)$ time, where $n$ is the number of variables.
Note that here we are making use of the incremental capabilities of the
solver, and reusing all asserted constraints, just doing one
additional \Verb"assert" per variable. To see the process in action,
consider the constraint \Verb"(2 * x) ~ 16".  The steps that we'd perform
are:
\begin{Verbatim}
(assert (= (* 2 x) 16))
(check-sat)
; SMT response: SAT
(get-value (x))
; SMT response: (8)
(push 1)
(assert (not (= x 8)))
(check-sat)
; SMT response: UNSAT
(pop 1)
; We generate a computed improvement: x ~ 8
\end{Verbatim}

\paragraph{Improve to Variable.}
We may also use the satisfying assignment to look for improvements of
the form $x = y$, where $x$ and $y$ are both variables.
We look at the satisfying assignment, ignoring variables
that were already improved to constants, and consider pairs of variables
that happen to have the same value in this assignment. If $x$ and $y$
are two such variables, then we try to prove that $x = y$ must hold
under the current assumptions:
\begin{Verbatim}
mustEqual s x y =
  solverProve s (eq (SMT.const x) (SMT.const y))
\end{Verbatim}
This process take $O(n^2)$ time, where $n$ is the number of variables,
but we only need to call the solver for pairs of variables that have the same
value---if they do not, then the current assignment provides a counter
example indicating that the variables do not need to be equal in general.

@@

The process of solving constraints is a straight-forward call the
solver.  The only thing we need to do before solving constraints is
to remove the assertions that were added during the consistency and
improvement stages.  Recall that GHC solves implication constraints
with two call to the constraint solver: the first one asserts
the given constraints, while the second one actually solves goals.

While we are asserting the assumption, we just check for consistency
and improvement (to generate new given equalities), but we perform
no solving, as there is nothing to solve.  During the solving stage,
we mark the solver's state after we've asserted the givens,
then we assert the new goals, check for consistency, and improvement,
and then, before solving, revert back to the state where only
the givens are asserted.
\begin{Verbatim}
solverSimplify s wanteds =
  solverPrepare s wanteds $ \others our_wanteds ->
  do res <- mapM tryToSolve our_wanteds
     let (unsolved, solved) = partitionEithers res
     return (solved, unsolved ++ others)
  where
  tryToSolve (ct,e) =
    do proved <- solverProve s e
       if proved
         then return (Right ct)
         else return (Left ct)
\end{Verbatim}
The call to \verb"solverPrepare" identifies the constraints that belong
to our theory and translates them the SMTLIB language, where \Verb"others"
are the constraints that are completely outside our theory, and
\Verb"out_wanteds" are the constraints that we know about. Then,
we invoke the solver for each goal, and see if we can prove it.

@@

\subsection{A Theory for Booleans}
We already made use of the lifted \Verb"Bool" type, but so far Haskell
does not support any interesting operations, beyond the \Verb"<=?" relation
that we described.  Programmers are already experimenting with defining
type functions. For example, with the current version of GHC we may
define a function for conjunction of type-level booleans:
\begin{Verbatim}
type family And a b where
  And False x = False
  And True x  = x
\end{Verbatim}

@@

GHC's constraint solver is not structured exactly in the same way as outlined
in the Nelson-Oppen simplification procedure, there are many similarities.
In particular, equalities play a central role in disseminating information,
but GHC currently does not do any back-tracking search (i.e., decision
procedure may only report inconsistency, or certain knowledge).  However,
there are situations where backtracking search is potentially useful.
For example, when solving class constraints in the context of overlapping
instances, or instance chains \cite{instance-chains}, it is useful
to consider the context of instances, but back-track and try a different
instance, if the context is found to be unsatisfiable.

GHC's implementation of Haskell contains numerous extensions, and so
there are many candidates for potential ``theories'', that is, constraints
that are solved by a dedicated decision procedure.  Engineering a
constraint solver in this style would be very beneficial, both
as one could reason about the correctness of each decision separately,
but also, because it would make it simpler to experiment with new
extensions to the system.

>>>>

For a few years now, there has been a steady push in the Haskell
community to explore and extend the type system, slowly approximating
functionality available in dependently typed languages
\cite{Eisenberg2012,Lindley2013,Eisenberg2014}.  The additional
expressiveness enables Haskell programmers to maintain more
invariants at compile time, which makes it easier to develop
reliable software, and also makes Haskell a nice language
for embedding domain specific languages \cite{ivory-experience}.
The Haskell compiler is not just a translator from source code
to executable binaries, but it also serves as a tool that
analyzes the program and helps find common mistakes early
in the development cycle.

@@

We illustrate the utility of the functionality provided by our
algorithm with a few short examples.  A very common example in
this area is to define a family of singleton types that links
type-level natural numbers to run-time constants that represent
them:
\begin{Verbatim}
data UNat :: Nat -> * where
  Zero :: UNat 0
  Succ :: UNat n -> UNat (n + 1)
\end{Verbatim}
Here we've used a unary representation of the natural numbers,
and each member of the family, \Verb"UNat n", has exactly one
inhabitant, namely the natural number \Verb"n" represented in
unary form.  Because we are using a GADT, we can pattern match
on the constructors of the type and gradually learn additional
information about the value being examined.  The kind \Verb"Nat"
is inhabited by types corresponding to the natural number
(e.g., \Verb"0", \Verb"1", \dots), and \Verb"(+)" is a type-level
function for adding natural numbers.

@@

The rest of the paper is organized as follows:
we start with a brief overview of SMT solvers form a user's perspective
(Section~\ref{smt}).  Then, in Section~\ref{GHC}, we introduce the
basic concepts of GHC's constraints
solver, which is necessary for putting the rest of the paper in context.
Section~\ref{algorithm} contains the details of the algorithm for
integrating an SMT solver with GHC, and Section~\ref{other} explains
other theories that could be added to GHC using the same technique.
Finally, Section~\ref{modular-typechecker} discusses the possibility
of using the ideas in this paper, and previous work on the
implementation of SMT solver, to engineer a modular constraint solver.

@@

isZero :: E a -> Bool
isZero (EInt x) = x == 0
\end{Verbatim}
When we check the definition of \Verb"isZero", we end up with an implication
constraint like this:
\begin{Verbatim}
(a ~ Int) => (Num a, Eq a)
\end{Verbatim}
Here, \Verb"a" is the type of the pattern variable \Verb"x", the wanted
constraints arise from the use of \Verb"0" and \Verb"(==)" respectively,
and the given constraint is obtained by pattern matching with \Verb"EInt"
because we know that the type parameter of \Verb"E" must be \Verb"Int".

@@

\subsection{Type-Checker Plugins}
In this paper we describe a fairly general extension to the constraint
solver, but other researchers are interested in different extensions,
for example, to add support for units of measure \cite{units-of-measure}.
Instead of having many
ad-hoc extensions directly
in GHC's constraint solver, we collaborated to define and implement an
API for extending GHC's functionality via {\em type-checker plug-ins}.
At present, we are aware of two main
users of this API---our work, and the work on units of measure. We
hope, however, that this mechanism would be useful to other researchers too,
as it makes it fairly easy to experiment with various extensions to
GHC's constraint solver.

@@

Instead, derived equalities help with type inference, by guiding the
instantiation of {\em unification variables}. In general,
it is always sound to instantiate a unification variable with whatever
type we want: doing so may reject valid programs, but it will not accept
invalid ones, because we still need to solve all necessary constraints.
Of course, we don't want to reject valid programs, and this is where derived
equalities help us: since they are implied by the goals and assumptions
together, we know that we are not loosing any generality when instantiating
as suggested by a derived constraint.  So, for example, if we compute a
derived constraint \Verb"x ~ Int", and we have a wanted constraint
\Verb"Eq x", and \Verb"x" is a unification variable, then we may
rewrite \Verb"Eq x" to \Verb"Eq Int", which we would proceed to solve
as usual.

@@

\paragraph{Input.} Currently, the algorithm is implemented as a type-checker
plug-in, and so the input to the algorithm is a collection of constraints
that GHC has determined to be inert. This means that GHC simplified everything
as much as it could, improved using equalities, and cannot see
anything else to do.  The inert constraints are presented to the plug-in
in three collections based on the constraints' provenance: one group contains
the given constraints (i.e., assumptions), one group contains derived
constraints (see Section~\ref{GHC} for details), and one group contains
the wanted constraints, which are the goals that need solving.

@@

\subsection{The Language of Constraints}
Our first task is to identify constraints that are relevant to our solver.
In the current implementation, we consider equalities and inequalities
between a subset of the type-expressions of kind \Verb"Nat".
The kind \Verb"Nat" is
inhabited by an infinite family of type-constants:
\begin{Verbatim}
0, 1, 2, .. :: Nat
\end{Verbatim}
These constants may be combined and compared using the following
type-level functions:
\begin{Verbatim}
type family (+)   :: Nat -> Nat -> Nat
type family (*)   :: Nat -> Nat -> Nat
type family (<=?) :: Nat -> Nat -> Bool
\end{Verbatim}
These functions have no user-specified definitions: instead, we extended
the core GHC simplifier with support for forward evaluation on concrete
values, so it will evaluate concrete expressions, such as \Verb"2 + 3".
This simple forward evaluation cleans up the constraints, leaving the more
complex reasoning---involving variables---to the algorithm being presently
described.  Technically, the clean-up is not necessary because the
plug-in will perform as much evaluation as it needs to solve the constraints,
but it is a lot more efficient to simply evaluate known constants without
consulting the external solver.  Also, at present there is no standard
way to declare that these type function are handled by a special solver---%
we could use a closed type family with no equations to prohibit user
defined instances, but this is a bit of an abuse of an unrelated feature
of the type system.

@@

The constraints supported by the external solver are summarized by the
following grammar:
\begin{align*}
c & = e \overset {\mathbb{N}} \sim e ~|~  e \overset {\mathbb{B}} \sim e \\
e^\mathbb{N} & = \alpha ~|~ n ~|~ e + e ~|~ n * e \\
e^\mathbb{B} & = \alpha ~|~ \mathrm{False} ~|~ \mathrm{True} ~|~ e \leq^? ~e \\
n          & = 0 ~|~ 1 ~|~ \dots
\end{align*}

@@

\paragraph{Natural Numbers.}
We are interested in working with natural numbers, however, the SMT
decision procedure works with integers.  To work around this mismatch,
we are careful that whenever we declare an SMT variable corresponding
to a type of kind \Verb"Nat", we also add a formula asserting that
the variable is not negative.

@@

The code samples in the rest of this Section use the following
API for interaction with the solver:
\begin{Verbatim}
data Solver :: *
data Expr   :: *
data Result = Sat | Unsat

@@

To check for consistency, we assert all constraints, and ask the SMT
solver if the result is satisfiable.  If this is not the case, then
we report and error, otherwise we proceed with the algorithm.
\begin{Verbatim}
checkConsistent :: Solver -> [Expr] -> IO ()
checkConsistent s cs =
  do mapM_ (solverAssert s) cs
     res <- solverCheck s
     return (res == Sat)
\end{Verbatim}

@@

Unfortunately, not all SMT solvers support this feature and CVC4---%
the SMT solver that we used during the development of this work---%
does not have this ability.  To work around this, we implemented
a custom minimization algorithm which work as follows:
\begin{Verbatim}
minimize :: Solver ->
            [Expr] -> -- part of conflict
            [Expr] -> -- search for conflict
            IO ()
minimize s yes [] = return yes
minimize s yes mb =
  do solverPush s
     search s yes [] mb

minimize :: Solver ->
            [Expr] -> -- part of conflict
            [Expr] -> -- currently asserted
            [Expr] -> -- search for conflict
            IO ()
search s yes mb (c : cs) =
  do solverAssert s c
     res <- solverCheck s
     case res of
       Unsat ->
         do solverPop s
            solverAssert s c
            minimize s (c : yes) mb

@@

\paragraph{Improve to Constant.} As we know that the constraints are
consistent, we can ask the SMT solver for a satisfying assignment.
This assignment contains one possible solution to constraints, but
we should not emit an improving equality unless we are sure that
this is the {\em the only} possible solution.  So, if $x = v$ is
in the satisfying assumption, then we try to prove that the currently
asserted constraints imply this fact.  We do this by temporarily asserting
$x \neq v$, and checking for satisfiability.  If the resulting is unsatisfiable,
then we know that $v$ is the only possible value for $x$,
and we can emit the corresponding improving equation.
\begin{Verbatim}
solverProve :: Solver -> Expr -> IO Bool
solverProve s p =
  do solverPush s
     solverAssert s (SMT.not p)
     res <- solverCheck s
     solverPop s
     return (res == Unsat)

mustBeK :: Solver -> Name -> Value -> IO Bool
mustBeK s x v =
  solverProve s (eq (SMT.const x) (value v))
\end{Verbatim}
This process takes $O(n)$ time, where $n$ is the number of variables.
Note that here we are making use of the incremental capabilities of the
solver, and reusing all asserted constraints, just doing one
additional \Verb"assert" per variable. To see the process in action,
consider the constraint \Verb"(2 * x) ~ 16".  The steps that we'd perform
are:
\begin{Verbatim}
(assert (= (* 2 x) 16))
(check-sat)
; SMT response: SAT
(get-value (x))
; SMT response: (8)
(push 1)
(assert (not (= x 8)))
(check-sat)
; SMT response: UNSAT
(pop 1)
; We generate a computed improvement: x ~ 8
\end{Verbatim}

\paragraph{Improve to Variable.}
We may also use the satisfying assignment to look for improvements of
the form $x = y$, where $x$ and $y$ are both variables.
We look at the satisfying assignment, ignoring variables
that were already improved to constants, and consider pairs of variables
that happen to have the same value in this assignment. If $x$ and $y$
are two such variables, then we try to prove that $x = y$ must hold
under the current assumptions:
\begin{Verbatim}
mustEqual :: Solver -> Name -> Name -> IO Bool
mustEqual s x y =
  solverProve s (eq (SMT.const x) (SMT.const y))
\end{Verbatim}
This process take $O(n^2)$ time, where $n$ is the number of variables,
but we only need to call the solver for pairs of variables that have the same
value---if they do not, then the current assignment provides a counter
example indicating that the variables do not need to be equal in general.

@@

The process of solving constraints is a straight-forward call to the
solver.  The only thing we need to do before solving constraints is
to remove the assertions that were added during the consistency and
improvement stages.  Recall that GHC solves implication constraints
with two calls to the constraint solver: the first one asserts
the given constraints, while the second one actually solves goals.

While we are asserting the assumption, we just check for consistency
and improvement (to generate new given equalities), but we perform
no solving as there is nothing to solve.  During the solving stage,
we mark the solver's state after we've asserted the givens,
then we assert the new goals, check for consistency, and improvement,
and then, before solving, revert back to the state where only
the givens are asserted.
\begin{Verbatim}
solverSimplify :: Solver ->
                  [Expr] -> IO ([Expr],[Expr])
solverSimplify s wanteds =
  solverPrepare s wanteds $ \others our_wanteds ->
  do res <- mapM tryToSolve our_wanteds
     let (unsolved, solved) = partitionEithers res
     return (solved, unsolved ++ others)
  where
  tryToSolve (ct,e) =
    do proved <- solverProve s e
       if proved
         then return (Right ct)
         else return (Left ct)
\end{Verbatim}
The call to \verb"solverPrepare" identifies the constraints that belong
to our theory and translates them the SMTLIB language, where \Verb"others"
are the constraints that are completely outside our theory, and
\Verb"out_wanteds" are the constraints that we know about. Then,
we invoke the solver for each goal, and see if we can prove it.

@@

\subsection{A Theory for Booleans}
We already made use of the lifted \Verb"Bool" type, but so far Haskell
does not support any interesting operations beyond the \Verb"<=?" relation
that we described.  Programmers are already experimenting with defining
type functions. For example, with the current version of GHC we may
define a function for conjunction of type-level booleans:
\begin{Verbatim}
type family And a b where
  And False x = False
  And True x  = x
\end{Verbatim}

@@

While GHC's constraint solver currently is structured quite differently
from an SMT solver, the similarities---in particular, the use of equality
constraints to disseminating information---suggest that there may be
a more modular way to structuring the solver.

While GHC currently does not do any back-tracking search (i.e., decision
procedure may only report inconsistency, or certain knowledge), this
is also the case for Nelson-Open if all theories are convex.  In addition,
there are interesting cases where backtracking might be potentially useful
in Haskell too.  For example, when solving class constraints in the
context of overlapping instances, or instance chains \cite{instance-chains},
it is useful to consider the context of instances,
but back-track and try a different instance, if the context is found
to be unsatisfiable.

GHC's implementation of Haskell contains numerous extensions, and so
there are many candidates for potential ``theories'', that is, constraints
that are solved by a dedicated decision procedure.  Engineering a
constraint solver in this style would be very beneficial, both
as one could reason about the correctness of each decision procedure
separately, but also, because it would make it simpler to experiment
with new extensions to the system.
